# %% [code]
---
title: "Podcast Listening Time"
date: '`r Sys.Date()`'
output:
  html_document:
    number_sections: true
    fig_caption: true
    toc: true
    fig_width: 7
    fig_height: 4.5
    theme: cosmo
    highlight: tango
    code_folding: hide
---
  
# Introduction  {.tabset .tabset-fade .tabset-pills}

The goal of this competition is to predict podcast listening time.

My notebook serves as a demonstration of some of the possible techniques available to arrive at a solution.  I intend to add to this as I have time available. Your questions and comments are welcome.

Lets dive right in.

The Kaggle kernels have many of the common r packages built in.  

## Load libraries

```{r }
#| label: setup
#| warning: false
#| message: false

if (dir.exists("/kaggle")){
  path <- "/kaggle/input/playground-series-s5e4/"

# options(repos = c(CRAN = "https://packagemanager.posit.co/cran/__linux__/focal/2021-05-17"))
# install.packages("ranger", quiet = TRUE)
    
options(repos = c(CRAN = "https://packagemanager.posit.co/cran/2021-03-22"))
#remotes::install_github("tidymodels/plsmod", quiet = TRUE)    
#remotes::install_github("mayer79/outForest", quiet = TRUE)    
#remotes::install_github("tidyverts/fable.binary", quiet = TRUE)
    
cores <- future::availableCores()

} else {
  path <- stringr::str_c(here::here("data"),"/")
  orig_path <- stringr::str_c(here::here("data"),"/")

  cores <- future::availableCores(omit = 2)
}
 
suppressPackageStartupMessages({
library(tidyverse, quietly = TRUE) # metapackage of all tidyverse packages
library(tidymodels) # metapackage see https://www.tidymodels.org/

library(correlationfunnel)    
   
library(bonsai)  

library(DALEXtra) # Tidymodels Explainability         
    
library(stacks)   # model ensembling

})

tidymodels_prefer()
conflicted::conflicts_prefer(brulee::coef)

options(tidymodels.dark = TRUE)

theme_set(ggdark::dark_theme_minimal())

```

## Load Data


```{r }
#| label: load data
#| warning: false
#| message: false




preprocessor <- function(dataframe) {

dataframe <- dataframe %>%
    janitor::clean_names() |>
  
    mutate(episode_title = parse_number(episode_title)) |> 

    mutate(number_of_ads = if_else(number_of_ads > 3 | is.na(number_of_ads), 4, number_of_ads)) |> 
  
    mutate(episode_length_minutes = if_else(episode_length_minutes > 121, 121, episode_length_minutes)) |> 

    mutate(across(c(where(is.character)), \(x) as.factor(x))) 

return(dataframe)
}

raw_df <- read_csv(str_c(path, "train.csv"),
                   show_col_types = FALSE) |> 
          preprocessor() 

tst_df <- read_csv(str_c(path, "test.csv"),
                   show_col_types = FALSE)  |>  
  preprocessor() 

# because we already know the test set, let's remove the train set factor levels that do not correspond with anything on the test set
for (col in names(raw_df)) {
    if (is.factor(raw_df[[col]]) & col != "listening_time_minutes") {
      # Get levels in train and test dataframes
      raw_levels <- levels(raw_df[[col]])
      tst_levels <- levels(tst_df[[col]])
      
      # Identify levels in train not in test
      new_levels <- setdiff(raw_levels, tst_levels)
      
      # Set these levels to NA in train dataframe
      raw_df[[col]] <- factor(raw_df[[col]], levels = c(tst_levels, new_levels))
      raw_df[[col]][raw_df[[col]] %in% new_levels] <- NA_character_
    }
  }

# the synthetic playground competitions seem to perform better when numerics are also included as factors
all_df <-
    bind_rows(raw_df %>% mutate(source = "train"),
              tst_df %>% mutate(source = "test")) 


train_df <- all_df %>% 
  filter(source == "train") %>% 
  select(-source) 

competition_df <- all_df %>% 
  filter(source == "test") %>% 
  select(-source, -listening_time_minutes)


```


# EDA {.tabset .tabset-fade .tabset-pills}

## Features

```{r}
#| label: Features
#| warning: false
#| message: false
#| fig.width: 6


features <- train_df %>%
  select(-id, -listening_time_minutes) |> 
  names()

train_df <- train_df %>% 
  distinct(pick(all_of(features)), .keep_all = TRUE)

nom_features <- train_df %>%
  select(all_of(features)) %>%
  select(where(is.character), where(is.factor)) %>%
  names() 

logical_features <- train_df %>%
  select(all_of(features)) %>%
  select(where(is.logical)) %>%
  names() 

num_features <- train_df %>%
  select(all_of(features)) %>%
  select(where(is.numeric)) %>%
  names()
```

Nominal features:

`r nom_features`

Numeric features: 

`r num_features`

Logical features: 

`r logical_features`


Size of the combined train and competition datasets:

`r nrow(all_df)`

Size of the split made available to machine learning

`r nrow(train_df)`



## Numeric features

```{r}
#| label: numeric
#| warning: false
#| message: false
#| fig.height: 6
#| fig.width: 6

train_df %>% 
  select(all_of(num_features), listening_time_minutes) %>% 
  pivot_longer(-listening_time_minutes,
    names_to = "metric",
    values_to = "value"
  ) %>%
  ggplot(aes(value)) +
  geom_histogram(aes(fill =  ggplot2::cut_number(listening_time_minutes,5)), position = "identity", bins = 30) +
  geom_freqpoly(aes(linetype = ggplot2::cut_number(listening_time_minutes,5)), bins = 30) +
  geom_rug(aes(color = ggplot2::cut_number(listening_time_minutes,5))) +
  facet_wrap(vars(metric), scales = "free", ncol = 2) +
  guides(color = "none") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "top",
        plot.title.position = "plot") +
  labs(color = NULL, 
       title = "Numeric Feature Univariate Distributions",
       caption = "Data: Kaggle.com | Visual: Jim Gruman")


```

## Nominal features

Explore the distribution of outcome class by factor level.

There appears to be some sort of interaction between genre and podcast_name. 


```{r}
#| label: nominal
#| warning: false
#| message: false
#| fig.height: 12
#| fig.width: 12


if(length(nom_features) >0){

train_df %>% 
  select(all_of(nom_features), listening_time_minutes) %>% 
  mutate(across(nom_features, fct_lump_n,n = 10, other_level = 'other')) %>%
  pivot_longer(-listening_time_minutes,
    names_to = "metric",
    values_to = "value"
  ) %>%
    
  filter(!is.na(listening_time_minutes)) %>% 
    
  summarise(n = n(),
            .by = c(listening_time_minutes, metric, value)) %>%
      
  mutate(value = tidytext::reorder_within(value, n, metric)) %>%
    
  ggplot(aes(x = n, y = value, fill = ggplot2::cut_number(listening_time_minutes, 5))) +
  geom_col() +
  tidytext::scale_y_reordered() +
  scale_x_continuous(n.breaks = 3, guide = guide_axis(n.dodge = 2))  +
  facet_wrap(vars(metric), scales = "free", ncol = 2) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
       legend.position = "bottom") +
  labs(title = "Nominal Feature Counts",
       fill = NULL,
       caption = "Data: Kaggle | Visual: Jim Gruman")

} 

train_df |> 
  summarise(mean_listening_time = mean(listening_time_minutes),
    .by = c(podcast_name, genre)) |> 
  ggplot(aes(genre, podcast_name, fill = mean_listening_time)) +
    geom_tile() +
  scale_fill_distiller(limits = c(0,100))


train_df |> 
  summarise(mean_listening_time = mean(listening_time_minutes),
    .by = c(podcast_name, publication_day)) |> 
  ggplot(aes(publication_day, podcast_name, fill = mean_listening_time)) +
    geom_tile() +
  scale_fill_distiller(limits = c(0,100))

train_df |> 
  summarise(mean_listening_time = mean(listening_time_minutes),
    .by = c(podcast_name, publication_time)) |> 
  ggplot(aes(publication_time, podcast_name, fill = mean_listening_time)) +
    geom_tile() +
  scale_fill_distiller(limits = c(0,100))

train_df |> 
  summarise(mean_listening_time = mean(listening_time_minutes),
    .by = c(episode_sentiment, podcast_name)) |> 
  ggplot(aes(episode_sentiment, podcast_name, fill = mean_listening_time)) +
    geom_tile() +
  scale_fill_distiller(limits = c(0,100))


```


## Counts of Missingness

The test set is missing one wind direction. We will fill it in with the median of all of the data.
                  
```{r}
#| label: counts of missingness

train_df %>% 
  summarize(across(all_of(features), function(x) sum(is.na(x)))) %>% 
  pivot_longer(everything(),
              names_to = "feature",
              values_to = "Count of Missing") %>% 
                   knitr::kable()

competition_df %>% 
  summarize(across(all_of(features), function(x) sum(is.na(x)))) %>% 
  pivot_longer(everything(),
              names_to = "feature",
              values_to = "Count of Missing") %>% 
                   knitr::kable()

```

## Counts of Distinct
                   
               
```{r}
#| label: counts of distinct
               
train_df %>%
  summarize(across(all_of(features), n_distinct)) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "Count of distinct train") |>
  left_join(
    competition_df %>%
      summarize(across(all_of(features), n_distinct)) %>%
      pivot_longer(everything(), names_to = "feature", values_to = "Count of distinct test"),
    by = join_by(feature)
  ) %>% 
                   knitr::kable()
               
```

## Duplicated

Is this competition transaction already in the training data with a correct listening_time_minutes?

```{r}
#| label: duplicates
#| warning: false
#| message: false

bind_rows(train_df %>% mutate(source = "train"),
              competition_df %>% mutate(listening_time_minutes= NA, source = "test")) |> 
    group_by_at(features) %>%
    mutate(num_dups = n(),
           dup_id = row_number()) %>% 
    ungroup() %>%
    group_by(source) %>%
    mutate(is_duplicated = dup_id > 1) %>% 
    count(is_duplicated) %>% 
                   knitr::kable()
               

```

## Pairwise Correlations
                   
`ggcorrplot` provides a quick look at numeric features where the correlation may be significant. 

```{r}
#| label: pairwise correlations
#| fig.width: 6
#| fig.height: 6                   
                   
train_df %>% 
  select(all_of(num_features), listening_time_minutes) %>%
  ggcorrplot::cor_pmat() %>% 
  ggcorrplot::ggcorrplot(hc.order = TRUE, lab = TRUE,
    type = "lower", insig = "blank") +
  labs(title = "Pairwise Correlations Training Set")
                   
competition_df %>% 
  select(all_of(num_features)) %>% 
  ggcorrplot::cor_pmat() %>% 
  ggcorrplot::ggcorrplot(hc.order = TRUE, lab = TRUE,
    type = "lower", insig = "blank") +
  labs(title = "Pairwise Correlations Competition Set")

``` 

## Correlation Funnel

Where the record is not missing, the episode length will drive the listening time.


```{r}
#| label: correlation funnel 
#| warning: false
#| message: false
#| fig.width: 6
#| fig.height: 6

train_df %>% 
  select(all_of(features)) %>%
    mutate(interact_genre_podcast_name = paste0(genre,"_",podcast_name)) |>                
    na.omit() |> 
    binarize(one_hot = TRUE) %>%
    bind_cols(train_df |> na.omit() |> select(listening_time_minutes)) |> 
    correlate(target = listening_time_minutes ) %>% 
    plot_correlation_funnel()

```                   
               

## Target

The outcome variable ranges from zero to 119.97 within the training set.

```{r}
#| label: outcome 
#| warning: false
#| message: false
#| fig.width: 6


train_df %>% 
  ggplot(aes(listening_time_minutes)) +
  geom_histogram(bins = 100) +
  labs(title = "Listening Time Minutes",
       caption = "Data: Kaggle.com | Visual: Jim Gruman")

```                            
           
                
# Machine Learning {.tabset .tabset-fade .tabset-pills}

## Recipe

```{r}
#| label: recipe

base_rec <- recipe(
    
    formula(paste0("listening_time_minutes ~ ", 
               str_c(features,  collapse = " + "))),
    data = train_df
  )  

                   

dummy_rec <- base_rec |> 

  step_mutate(interact_genre_podcast_name = paste0(genre,"_",podcast_name)) |>
  
  embed::step_lencode_glm(interact_genre_podcast_name, outcome = vars(listening_time_minutes)) |> 
   
  step_impute_linear(episode_length_minutes,
                  impute_with = imp_vars(interact_genre_podcast_name, publication_day, publication_time, number_of_ads)) |> 

  step_impute_linear(guest_popularity_percentage,
                  impute_with = imp_vars(interact_genre_podcast_name, publication_day, publication_time, number_of_ads)) |> 

  step_dummy(all_nominal_predictors()) |> 
                     
  step_zv(all_predictors()) %>%
  step_normalize(all_predictors()) 


folds <- rsample::vfold_cv(train_df,
                           v = 5)


```



## Workflowset Ensemble

```{r}
#| label: workflowset
#| warning: false
#| message: false
#| fig.width: 24


linear_reg_glmnet_spec <-
  linear_reg(penalty = tune(), mixture = 1) %>%
  set_engine('glmnet')

boost_tree_xgb_spec <- 
  boost_tree(
    trees = 120L,
    min_n = tune()
  ) %>% 
  set_engine(engine = "xgboost", nthread = cores) %>%
  set_mode(mode = "regression")                     

boost_tree_lgbm_spec <- 
  boost_tree(
    trees = 120L,
   tree_depth = tune(),
   learn_rate =  tune(),
   min_n = tune(),
   loss_reduction = 0
  ) %>% 
  set_engine(engine = "lightgbm",
             is_unbalance = TRUE,
             num_leaves = tune(),
             num_threads = cores
             ) %>%
  set_mode(mode = "regression") 
                   

dep_models <- 
   workflow_set(
      preproc = list(base = base_rec,
                     dummy = dummy_rec,
                     dummy = dummy_rec),
      models = list(lgbm = boost_tree_lgbm_spec,
                    xgb = boost_tree_xgb_spec,
                    linear = linear_reg_glmnet_spec),
      cross = FALSE
   ) %>% 
  option_add_parameters() |> 
  option_add(
    control = finetune::control_sim_anneal(save_pred = TRUE,save_workflow = TRUE, verbose = TRUE, verbose_iter = TRUE),
    metrics = metric_set(rmse)
  )

glm_params <- dep_models |> 
    extract_workflow("dummy_linear") |> 
  parameters() |> 
  update(
#    mixture = mixture(range = c(0.7,1)),
    penalty = dials::penalty(range = c(-4,-2)))

xgb_params <- dep_models |>
       extract_workflow("dummy_xgb") |>       
  parameters() |> 
  update(
    min_n = min_n(range = c(3,100)))       
                   
lgbm_params <- dep_models |>
       extract_workflow("base_lgbm") |>       
  parameters() |> 
  update(
    tree_depth = tree_depth(range = c(5,100)),
    learn_rate = learn_rate(range = c(-1.3,-0.8)),
    min_n = min_n(range = c(3,100)),
    num_leaves = num_leaves(range = c(5,100)))       
                   
dep_models <- dep_models |> 
  option_add(
    param_info = glm_params,
    id = "dummy_linear"
  ) |> 
  option_add(
    param_info = xgb_params,
    id = "dummy_xgb"
  ) |> 
  option_add(
    param_info = lgbm_params,
    id = "base_lgbm"
  ) |> 
   workflow_map("tune_sim_anneal", resamples = folds, iter = 10, 
                metrics = metric_set(rmse), verbose = TRUE)

rank_results(dep_models, select_best = TRUE)                    

autoplot(dep_models) +
  geom_text(aes(y = mean -0.03, label= wflow_id), angle = 90, hjust = 1)+
  theme(legend.position = "none")

```

## Hyperparameters and Feature Importance

```{r }                   
#| label: linear Details
#| warning: false
#| message: false                
                   

dep_models %>%
  dplyr::filter(grepl("dummy_linear", wflow_id)) %>%
  dplyr::mutate(metrics = purrr::map(result, tune::collect_metrics)) %>%
  dplyr::select(wflow_id, metrics) %>%
  tidyr::unnest(cols = metrics) |> 
  arrange(mean)

dep_models |> 
  workflowsets::extract_workflow_set_result("dummy_linear") |> 
  autoplot() +
  labs(title = "GLMNET Hyperparameter Search")

best_linear_params <- dep_models |> 
  workflowsets::extract_workflow_set_result("dummy_linear") |> 
  select_best(metric = "rmse")  

dep_models |> 
  workflowsets::extract_workflow("dummy_linear") |> 
  finalize_workflow(best_linear_params) |> 
  fit(train_df) |> 
  vip::vip(num_features = 50L)                   
                                
                   
                   
```

```{r }                   
#| label: xgb Details
#| warning: false
#| message: false                
                   

dep_models %>%
  dplyr::filter(grepl("dummy_xgb", wflow_id)) %>%
  dplyr::mutate(metrics = purrr::map(result, tune::collect_metrics)) %>%
  dplyr::select(wflow_id, metrics) %>%
  tidyr::unnest(cols = metrics) |> 
  arrange(mean)

dep_models |> 
  workflowsets::extract_workflow_set_result("dummy_xgb") |> 
  autoplot() +
  labs(title = "XGBoost Hyperparameter Search")

best_linear_params <- dep_models |> 
  workflowsets::extract_workflow_set_result("dummy_xgb") |> 
  select_best(metric = "rmse")  

dep_models |> 
  workflowsets::extract_workflow("dummy_xgb") |> 
  finalize_workflow(best_linear_params) |> 
  fit(train_df) |> 
  vip::vip(num_features = 50L)                   
                                
                   
                   
```                   


```{r }                   
#| label: lgbm Details
#| warning: false
#| message: false 
                   
dep_models %>%
  dplyr::filter(grepl("base_lgbm", wflow_id)) %>%
  dplyr::mutate(metrics = purrr::map(result, tune::collect_metrics)) %>%
  dplyr::select(wflow_id, metrics) %>%
  tidyr::unnest(cols = metrics) |> 
  arrange(mean)

dep_models |> 
  workflowsets::extract_workflow_set_result("base_lgbm") |> 
  autoplot() +
  labs(title = "LightGBM Hyperparameter Search")   

best_lgbm_params <- dep_models |> 
  workflowsets::extract_workflow_set_result("base_lgbm") |> 
  select_best(metric = "rmse")  

dep_models |> 
  workflowsets::extract_workflow("base_lgbm") |> 
  finalize_workflow(best_lgbm_params) |> 
  fit(train_df) |> 
  vip::vip(num_features = 50L)   


                   
```           

## Ensemble            

                   
```{r }                   
#| label: Ensemble Fit
#| warning: false
#| message: false 

dep_stack <- stacks() %>%
  add_candidates(dep_models) %>%
  blend_predictions(  metric = metric_set(rmse),
      penalty = c(10^seq(-2.5, -0.6, 0.1)),
      non_negative = TRUE,
      control = tune::control_grid(allow_par = TRUE))

autoplot(dep_stack)

autoplot(dep_stack, type = "members")  
autoplot(dep_stack, type = "weights")
autoplot(dep_stack, type = "performance")

regression_fit <- dep_stack %>% 
    fit_members()

```

# Performance {.tabset .tabset-fade .tabset-pills}


## Residuals

```{r}
#| label: residuals

augment(regression_fit, train_df) |>
    slice_sample(prop = 0.01 ) |>
    ggplot(aes(listening_time_minutes, listening_time_minutes - .pred)) +
    geom_point(alpha = 0.2) +
    geom_smooth(se = FALSE) +
    labs(title = "Listening Time Minutes",
       caption = "Data: Kaggle.com | Visual: Jim Gruman")

```                   

## DALEX

```{r}
#| label: explainer
#| warning: false
#| message: false


explainer <- 
  explain_tidymodels(
    regression_fit, 
    data = train_df %>% dplyr::select(all_of(features)), 
    y = train_df$listening_time_minutes,
    label = "Ensemble",
    verbose = FALSE
  )  %>% 
  model_parts()

ggplot_imp <- function(...) {
  obj <- list(...)
  metric_name <- attr(obj[[1]], "loss_name")
  metric_lab <- paste(metric_name, 
                      "after permutations\n(higher indicates more important)")
  
  full_vip <- bind_rows(obj) %>%
    filter(variable != "_baseline_")
  
  perm_vals <- full_vip %>% 
    filter(variable == "_full_model_") %>% 
    group_by(label) %>% 
    summarise(dropout_loss = mean(dropout_loss))
  
  p <- full_vip %>%
    filter(variable != "_full_model_") %>% 
    mutate(variable = fct_reorder(variable, -dropout_loss)) %>%
    ggplot(aes(dropout_loss, variable)) 
  
  if(length(obj) > 1) {
    p <- p + 
      facet_wrap(vars(label)) +
      geom_vline(data = perm_vals, aes(xintercept = dropout_loss, color = label),
                 linewidth = 1.4, lty = 2, alpha = 0.7) +
      geom_boxplot(aes(color = label, fill = label), alpha = 0.2)
  } else {
    p <- p + 
      geom_vline(data = perm_vals, aes(xintercept = dropout_loss),
                 linewidth = 1.4, lty = 2, alpha = 0.7) +
      geom_boxplot(fill = "#91CBD765", alpha = 0.4)
    
  }
  p +
    theme(legend.position = "none") +
    labs(x = metric_lab, 
         y = NULL,  fill = NULL,  color = NULL)
}
                   
ggplot_imp(explainer)      
                   

```                   

## Submission
```{r }                   
#| label: submission
#| warning: false
#| message: false
                   

submit_df <-  augment(regression_fit , competition_df) %>%
       transmute(id = round(id), listening_time_minutes = .pred)

head(submit_df)  %>% 
     knitr::kable()      

submit_df  %>% 
  write_csv("submission.csv")
```  